A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-06 tex/BiSig.tex
[ tex ] Use `\aritysymbol` in S3; explain the notation more carefully
[ tex ] Minor revisions up to S4

\usepackage{bbold}
\newcommand{\hmmax}{0}
\newcommand{\bmmax}{0}
\usepackage{bm}
\usepackage{mathtools}
\input{mathlig}
\usepackage{tikz}
\usetikzlibrary{matrix,arrows}
\usepackage{mathpartir}
\usepackage{xspace}
\usepackage[capitalise,noabbrev]{cleveref}

@@

\usepackage{microtype}
\microtypesetup{babel}
\usepackage{subfiles}

\citestyle{acmauthoryear}
\AtEndPreamble{%
\theoremstyle{acmdefinition}
\newtheorem{remark}[theorem]{Remark}
}

@@

\input{type-notation} %% for type theory notation copied from HoTT Book
\input{macros}
\usepackage{agda-lhs2tex}

@@

\begin{abstract}
%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
  First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%  which is analogous to a grammar in parsing.
  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also formulated in a way that unifies computation and proof, resulting in a compact and practical theory---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
  The main theorem is logically a trichotomy on parsed syntax trees and computationally a synthesiser that may throw an exception due to missing type annotations.
  By being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
\end{abstract}

@@

\begin{acks}
We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.

The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
\end{acks}

>>>>

\usepackage{subfiles}
\usepackage{bbold}
\newcommand{\hmmax}{0}
\newcommand{\bmmax}{0}
\usepackage{bm}
\usepackage{mathtools}
\input{mathlig}
\usepackage{tikz}
\usetikzlibrary{matrix,arrows}
\usepackage{mathpartir}
\usepackage{xspace}
\usepackage[capitalise,noabbrev]{cleveref}

% Used for typesetting Agda code
\usepackage{agda-lhs2tex}

@@

\usepackage{microtype}
\microtypesetup{babel}

@@

\input{type-notation} %% for type theory notation copied from HoTT Book
\input{macros}

\citestyle{acmauthoryear}

\AtEndPreamble{%
\theoremstyle{acmdefinition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem*{claim}{Claim}
\newtheorem{claimproof}{Proof of Claim}
}

@@

\begin{abstract}
%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
  First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%  which is analogous to a grammar in parsing.
  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
  The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
  By being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
\end{abstract}

@@

\begin{acks}
%We thank Nathanael Arkor for the useful conversation.
The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
\end{acks}

### 2023-07-06 tex/sec1-intro.tex
[ tex ] Minor revisions up to S4

Type inference is an important mechanism for the transition to well-typed programs from parsed abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain some type annotations.
\citet{Dunfield2021} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

@@

Within the theory, we formulate essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesiser and ensure that it implements the general specification, formalising what were only informally outlined by~\citet{Dunfield2021}.
We also pay attention to the problem of whether a raw term has enough type annotations, often overlooked in the literature, by introducing \emph{mode preprocessing}, which assigns a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
Practically, by running a mode preprocessor, the user can first make sure that a raw term has enough annotations, and then deal with typing issues without worrying about missing annotations anymore.
Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has an ordinary typing derivation.

>>>>

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain some type annotations.
\citet{Dunfield2021} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

@@

Within the theory, we formulate essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesiser and ensure that it implements the general specification, formalising what were only informally outlined by~\citet{Dunfield2021}.
We also pay attention to the problem of whether a raw term has enough type annotations, often overlooked in the literature, by introducing \emph{mode preprocessing}, which assigns a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
Practically, by running a mode preprocessor, the user can first make sure that a raw term has enough annotations, and then deal with typing issues without worrying about missing annotations anymore.
Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation.

### 2023-07-06 tex/sec2-outline.tex
[ tex ] Minor revisions up to S4

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
To give more precise definitions:
The raw terms%
\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
for simply typed $\lambda$-calculus are defined in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to help with type synthesis.
Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
Now we can define what it means to solve the type synthesis problem.

\begin{definition}
\label{def:STLC-type-synthesiser}
Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $|\Gamma| \vdash t$ (where $|\Gamma|$~is the list of variables appearing in~$\Gamma$) as input, and establishes one of the following outcomes:
\begin{enumerate}
\item there exists a derivation of $\Gamma \vdash t : A$ for some type~$A$,
\item there does not exist a derivation $\Gamma \vdash t : A$ for any type~$A$, or
\item $E$~holds for~$t$.
\end{enumerate}
\end{definition}

@@

Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.

\Josh{Here we briefly recap the basic ideas of bidirectional type synthesis.
If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.}

@@

\begin{figure}
  \centering
  \small
  \judgbox{V |- \isTerm{t}^\dir{d}}{A raw term~$\isTerm{t}$ is in mode~$d$ with free variables in $V$}
  \begin{mathpar}
    \inferrule{x \in V}{V |- \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{V |- \isTerm{t}^{\chk}}{V |- (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{V |- \isTerm{t}^{\syn}}{V |- \isTerm{t}^{\chk}}\,\ChkRule{Sub}
    \\
    \inferrule{V, x |- \isTerm{t}^{\chk}}
    {V |- (\isTerm{\lam{x}t})^{\chk}}\,\ChkRule{Abs}
    \and
    \inferrule{V |- \isTerm{t}^{\syn} \\ V |- \isTerm{u}^{\chk}}
    {V |- (\isTerm{t\;u})^{\syn}}\,\SynRule{App}
  \end{mathpar}
  \caption{Mode derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-mode-derivations}
\end{figure}

@@

\begin{definition}
\label{def:STLC-mode-preprocessor}
A \emph{mode preprocessor} decides for any raw term~$V \vdash \isTerm{t}$ and mode~$\dir{d}$ whether $V \vdash \isTerm{t}^\dir{d}$.
\end{definition}

One (less important) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so the term $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.

@@

\begin{definition}
\label{def:STLC-bidirectional-type-synthesiser}
A \emph{bidirectional type synthesiser} decides for any context~$\Gamma$ and synthesising term $|\Gamma| \vdash \isTerm{t}^{\syn}$ whether $\Gamma \vdash \isTerm{t} \syn A$ for some type~$A$.
\end{definition}

@@

\begin{figure}
  \small
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes, row sep=3em, column sep=10em]
  {
    \boxed{\erase{\Gamma} \vdash \isTerm{t}}   & \boxed{\erase{\Gamma} \vdash \isTerm{t}^{\syn}} \\
    \boxed{\Gamma \vdash \isTerm{t} : A} & \boxed{\Gamma \vdash \isTerm{t} \syn A} \\
  };
  \path[->, line width=0.5pt]
      (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)
      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2)
      (m-2-2) edge[bend right=8] node [above] {soundness}    (m-2-1);
  \path[dashed,->,line width=0.5pt]
      (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);
  \end{tikzpicture}
  \caption{An implementation of type synthesis via bidirectional typing}
\end{figure}

The construction is straightforward:
Run the mode preprocessor on the input term~$|\Gamma| \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $|\Gamma| \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound}%
with respect to the original typing~(\cref{fig:STLC-typing-derivations}).

@@

\begin{lemma}[Completeness]
If\/ $|\Gamma| \vdash t^\dir{\syn}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
\end{lemma}

We will construct a mode preprocessor~(\cref{sec:mode-preprocessing}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
To quantify over all such systems, we need their generic definitions, which we formulate in \cref{sec:defs}.

>>>>

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
To give more precise definitions:
The raw terms for simply typed $\lambda$-calculus are defined%
\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.
Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
Now we can define what it means to solve the type synthesis problem.

\begin{definition}
\label{def:STLC-type-synthesiser}
Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $\erase\Gamma \vdash t$ (where $\erase\Gamma$~is the domain of~$\Gamma$) as input, and establishes one of the following outcomes:
\begin{enumerate}
\item there exists a derivation of $\Gamma \vdash t : A$ for some type~$A$,
\item there does not exist a derivation $\Gamma \vdash t : A$ for any type~$A$, or
\item $E$~holds for~$t$.
\end{enumerate}
\end{definition}

@@

Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
It is often said that a type synthesis algorithm can be `read off' from well-designed bidirectional typing rules.
Take the $\SynRule{App}$ rule as an example:
to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the type $A \bto B$.
Note that the synthesiser is able to figure out the type~$A$ for checking~$u$ and the type~$B$ to be synthesised for $t\;u$ because they have been computed when synthesising the type $A \bto B$ of~$t$.
In general, there should be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised).
This is called \emph{mode-correctness}, which we will formally define in \cref{sec:mode-correctness}.

@@

\begin{figure}
  \centering
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{V |- \isTerm{t}^{\syn}}$ & A raw term~$\isTerm{t}$ (with free variables in~$V$) is in synthesising mode \\
    $\boxed{V |- \isTerm{t}^{\chk}}$ & A raw term~$\isTerm{t}$ (with free variables in~$V$) is in checking mode
  \end{tabular}
  \egroup
  \begin{mathpar}
    \inferrule{x \in V}{V |- \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{V |- \isTerm{t}^{\chk}}{V |- (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{V |- \isTerm{t}^{\syn}}{V |- \isTerm{t}^{\chk}}\,\ChkRule{Sub}
    \\
    \inferrule{V, x |- \isTerm{t}^{\chk}}
    {V |- (\isTerm{\lam{x}t})^{\chk}}\,\ChkRule{Abs}
    \and
    \inferrule{V |- \isTerm{t}^{\syn} \\ V |- \isTerm{u}^{\chk}}
    {V |- (\isTerm{t\;u})^{\syn}}\,\SynRule{App}
  \end{mathpar}
  \caption{Mode derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-mode-derivations}
\end{figure}

@@

\begin{definition}
\label{def:STLC-mode-preprocessor}
A \emph{mode preprocessor} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
\end{definition}

One (less important) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so a term of the form $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.

@@

\begin{definition}
\label{def:STLC-bidirectional-type-synthesiser}
A \emph{bidirectional type synthesiser} decides for any context~$\Gamma$ and synthesising term $\erase\Gamma \vdash \isTerm{t}^{\syn}$ whether $\Gamma \vdash \isTerm{t} \syn A$ for some type~$A$.
\end{definition}

@@

The construction is straightforward:
Run the mode preprocessor on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).

@@

\begin{lemma}[Completeness] \label{thm:STLC-completeness}
If\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
\end{lemma}

We will construct a mode preprocessor~(\cref{sec:mode-preprocessing}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
To quantify over all such systems, we need their generic definitions, which we formulate next.

%\begin{figure}
%  \small
%  \begin{tikzpicture}
%  \matrix (m) [matrix of math nodes, row sep=3em, column sep=10em]
%  {
%    \boxed{\erase{\Gamma} \vdash \isTerm{t}}   & \boxed{\erase{\Gamma} \vdash \isTerm{t}^{\syn}} \\
%    \boxed{\Gamma \vdash \isTerm{t} : A} & \boxed{\Gamma \vdash \isTerm{t} \syn A} \\
%  };
%  \path[->, line width=0.5pt]
%      (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
%      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)
%      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2)
%      (m-2-2) edge[bend right=8] node [above] {soundness}    (m-2-1);
%  \path[dashed,->,line width=0.5pt]
%      (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);
%  \end{tikzpicture}
%  \caption{An implementation of type synthesis via bidirectional typing}
%\end{figure}
